Categorical Foundations of E-Graphs With Binding

### 2025-07-10 POPL2026/paper/sections/combinatorial_semantics.tex
lots of polish

\section{String Diagram Rewrite Theory}\label{sec:combinatorial-semantics}

We recall the fundamental definitions and results of string diagram rewrite theory for symmetric monoidal categories,  recapitulating the correspondence between string diagram rewriting and an appropriate notion of double pushout (DPO) rewriting of certain hypergraphs,  as established in~\cite{bonchi_string_2022-1, bonchi_string_2022-2}.

First,  we fix some notation.  Let $(-)^*$ be the free monoid monad over \textsf{Set}.
We extend this to the case of relations $R \subseteq V \times V$,  so that we denote by $R^{*} \subseteq V^* \times V^*$ the element-wise extension of $R$ to a relation on ordered sequences.
We will sometimes use functional notation for relations,  so that given $R \subseteq V \times W$ and $v \in V$,  $R(v) \subseteq W$.
We elide associativity and unit isomorphisms associated with coproducts $(+)$,  and denote by $\iota_j: X_{j} \rightarrow X_{1} + \ldots X_{j} + \ldots X_{n}$ the $j^{\text{th}}$ injection,  and $[f,g]: A_1 + A_2 \to B$ the co-pairing of two morphisms $f: A_1 \to B$ and $g:A_2 \to B$.
We denote by $A +_{f,g} B$ the pushout of the span $A \xleftarrow{f} C \xrightarrow{g} B$.
We will also refer to $A,B$ as \textit{feet} of the cospan, and to $C$ as \textit{carrier}.
% Given an ordered sequence of labels from some alphabet $\Sigma$, $[l_1, \ldots, l_n]$ we will denote with $w([l_1, \ldots, l_n])$ the word concatenated from these labels using $\otimes$, \textit{e.g.}, $w([A,B]) = A \otimes B$.

@@

In ~\cite{bonchi_string_2022-2}, certain class of hypergraphs \textit{with interfaces} and their DPO rewriting are shown to be sound and complete for categorical presentations of arbitrary SMTs.
Hypergraphs generalise graphs by allowing edges to have multiple sources and targets.
When interpreting string diagrams as hypergraphs,  morphisms are interpreted as edges,  and wires as vertices.
However,  there is also a need to specify which vertices are to be considered as \textit{input} and \textit{outputs},  corresponding to the dangling wires at the bottom and top of a string diagram.
This gives rise to the definition of a hypergraph with interfaces.
% Subsequent work \cite{bonchi_string_2022-2} builds on this correspondence to give a restriction of hypergraphs with interfaces (called \textit{monogamous} directed acyclic),  leading to a correspondence between symmetric monoidal string diagrams (\textit{i.e.}, without requiring a Frobenius algebra) and the category of (cospans of) such hypergraphs.
We then evoke the definition of hypergraphs and corresponding cospans as introduced in~\cite{bonchi_string_2022-1,bonchi_string_2022-2}.

\begin{definition}[Category of hypergraphs]\label{def:hypergraph}
	A \emph{hypergraph $\mathcal{G}$ over monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$} is a tuple $(V,E,s,t,l)$,  where $V$ is a finite set of vertices, $E$ is a finite set of edges, $s : E \to V^{*}$ is a source function, $t : E \to V^{*}$ is a target function,  and $l = (l_{V} : V \to \Sigma_{O},l_{E} : E \to \Sigma_{M})$ is a pair of labelling functions that assigns each edge and vertex an operator and a type from monoidal signature $\Sigma$ respectively.
	The labelling function must respect the typing of the operator: for all edges $e$, such that $l(e) = c : A \to B$,  $l_{V}^{*}(s(e)) = A$ and $l_{V}^{*}(t(e)) = B$.
	We call a hypergraph \textit{discrete} if its set of edges is empty.
	A \emph{hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ is given by a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that the following hold.
	\begin{enumerate}
		\item $\phi_V^*(s_{\mathcal{F}}(e)) = s_{\mathcal{G}}(\phi_E(e)) \qquad \phi_V^*(t_{\mathcal{F}}(e)) = t_{\mathcal{G}}(\phi_E(e))$
		\item $l_{V,\mathcal{F}}(v) = l_{V,\mathcal{G}}(\phi_V(v)) \qquad\;\; l_{E,\mathcal{F}}(e) = l_{E,\mathcal{G}}(\phi_E(e))$
	\end{enumerate}
\end{definition}
We denote by $\catname{Hyp(\Sigma)}$ the category of hypergraphs over $\Sigma$ and hypergraph homomorphisms.
$\catname{Hyp(\Sigma)}$ has all finite colimits and, in particular,  the coproduct is given by the disjoint union of hypergraphs.
The initial object is given by the empty hypergraph.

To properly interpret morphisms of a symmetric monoidal category we need to account for the dangling wires at the diagram's top and bottom that represent inputs and outputs of a morphism.
Such dangling wires are represented as interfaces for a given hypergraph giving rise to a category of cospans of hypergraphs as defined next.

% \begin{definition}[Symmetric monoidal category of cospans]
% Let $\mathbb{C}$ be a category with all finite colimits.  
% A cospan from $X$ to $Y$ is a pair of arrows $X \xrightarrow{} A \xleftarrow{} Y$  in $\mathbb{C}$.  This \textit{category of cospans of $\mathbb{C}$},  denoted $\catname{Csp(\mathbb{C})}$,  has as objects those of $\mathbb{C}$ and as arrows isomorphism classes of cospans.

% Composition of cospans $X \xrightarrow{f} A \xleftarrow{g} Y$ and $Y \xrightarrow{h} B \xleftarrow{i} Z$ is given by pushout: $X \xrightarrow{f} A +_{g,h} B \xleftarrow{i} Z$,  with identity given by the cospan of identities.  Its monoidal product is given by the coproduct of $\mathbb{C}$ as follows: 
% \begin{multline*}
%     (A \xrightarrow{f} \mathcal{G} \xleftarrow{g} B) \otimes (A' \xrightarrow{f'} \mathcal{G'} \xleftarrow{g'} B') \coloneq\\ A + A' \xrightarrow{f+f'} \mathcal{G} + \mathcal{G'} \xleftarrow{g+g'} B + B'  , 
% \end{multline*}
% and its unit by the initial object of $\mathbb{C}$.
% Symmetry is inherited from $\mathbb{C}$ as the coproduct~\cite{MonoidalCoproduct}.
% \end{definition}

% Then, morphisms of $\catname{S}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces~\cite{bonchi_string_2022-2}.
\begin{definition}[Hypergraphs with interfaces]
	\label{def:cspd}
	A hypergraph with interfaces is a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\catname{Hyp}(\Sigma)$ where $n$, $m$ are discrete ordered hypergraphs that denote interfaces.
	Define the category of \emph{hypergraphs with interfaces} $\catname{HypI(\Sigma)}$ as the category with discrete \textit{ordered} hypergraphs as objects and isomorphism classes of hypergraphs with interfaces as morphisms.
\end{definition}
$\HypI{\Sigma}$ is a symmetric monoidal category with the tensor product given by coproduct in $\catname{Hyp}(\Sigma)$:
\begin{align*}
	n \xrightarrow{f}   & \mathcal{F} \xleftarrow{g} m                                                                                   \\
	                    & \times \qquad \quad \Coloneqq n + n' \xrightarrow{f + f'} \mathcal{F} + \mathcal{G} \xleftarrow{g + g'} m + m' \\
	n' \xrightarrow{f'} & \mathcal{G} \xleftarrow{g'} m'
\end{align*}
Unit is given by the initial object of $\catname{Hyp}(\Sigma)$.
Symmetry is inherited from coproduct in $\catname{Hyp}(\Sigma)$~\cite{MonoidalCoproduct}.

Composition of two cospans $n \xrightarrow{f} \mathcal{F} \xleftarrow{g} m;m \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} k$ is given by the pushout
$
	n \xrightarrow{f} \mathcal{F} +_{g,f'} \mathcal{G} \xleftarrow{g'} k
$,
computed by identifying vertices in $\mathcal{F} + \mathcal{G}$ that share pre-image in $m$.
As pushout is only defined up to isomorphism, it justifies the morphisms in this category being isomorphism classes of cospans.
To make such cospans complete with respect to representing morphisms of $\catname{S}(\Sigma)$, we need to impose some restrictions on their form.
\begin{definition}[Degree,  path,  acyclic]
	The \emph{in-degree} of a vertex $v$ in a hypergraph $\mathcal{G}$ is the number of hyperedges $e \in {E_\mathcal{G}}$ such that $v \in t(e)$.
	Similarly, the out-degree of $v$ is the number of edges $e \in E_\mathcal{G}$ such that $v \in s(e)$.
	A \emph{path} from $e_0$ to $e_{n-1}$ of length $n$ in a hypergraph $\mathcal{G}$ is an ordered sequence of edges $[e_0, e_1, \ldots, e_{n-1}] \in E_\mathcal{G}^*$ such that for all $i < n - 1$ some vertex $v \in t(e_i)$ is also in $s(e_{i+1})$.
	A hypergraph is \emph{directed acyclic} if it contains no paths of length $n > 0$ that start and end in the same edge.
\end{definition}

\begin{definition}[Category of MDA Hypergraphs with Interfaces~\cite{bonchi_string_2022-2}]
	\label{def:monogamy_hyp}
	We call a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\HypI{\Sigma}$ \emph{monogamous directed acyclic (MDA)} if:
	\begin{enumerate}
		\item $\mathcal{G}$ is a directed acyclic hypergraph;
		\item $f$ and $g$ are monomorphisms;
		\item the in-degree and out-degree of every vertex is at most~$1$;
		\item vertices with in-degree $0$ are precisely the image of $f$;
		\item vertices with out-degree $0$ are precisely the image of $g$.
	\end{enumerate}
	We denote by $\MdaCospans$ the subcategory of $\HypI{\Sigma}$ of MDA cospans.
\end{definition}
Note that $\MdaCospans$ remains a symmetric monoidal category with the same equipment as before.  We now recall from
\cite{bonchi_string_2022-2} the following theorem,  which says the category $\MdaCospans$ is in fact \textit{equivalent} to $\catname{S}(\Sigma)$.
\begin{theorem}[Corollary 26 of~\cite{bonchi_string_2022-2}]\label{thm:prop-equiv}
	$\catname{S}(\Sigma) \cong \MdaCospans$
\end{theorem}

The practical meaning of the above theorem is that two $\Sigma$-terms $f$ and $g$ are equal modulo SMC equations if and only if they are represented as the same hypergraph with interfaces.
The theorem defines a full and faithful functor that is induced by the interpretation function of morphism terms that renders the latter as cospans of hypergraphs and is given in Appendix~\ref{sec:appendix:interpretation}.
Term rewriting and hence equality modulo $\mathcal{E}$ is then implemented as DPO rewrites over such hypergraphs with interfaces that we recall next.

\subsection{DPOI-Rewriting for Hypergraphs with Interfaces}
DPO rewriting is a standard technique for formalising graph transformations categorically.
% In this section we recall standard results of double-pushout (DPO) rewriting of hypergraphs,  as well as the specific notion of DPO rewriting which is sound and complete for arbitrary SMTs,  which is called \textit{convex} rewriting.
The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $\mathcal L \rightsquigarrow \mathcal R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $\mathcal L$ within $\mathcal{G}$,  remove it,  and then plug $\mathcal R$ into the hole to recover the rewritten graph $\mathcal{H}$.  In DPO rewriting,  such a rewrite rule is formalised as a span $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$ in some appropriate category of graphs,  where $\mathcal K$ is some invariant that should be maintained during rewriting.
A typical DPO square is depicted in Fig.~\ref{fig:dpo_dpoi}~(left).

The morphism $\mathcal L \to \mathcal G$ is called a \textit{match},  defined whenever there is a corresponding \textit{pushout complement}---the pair of morphisms $\mathcal K \to \mathcal \mathcal{L}^{\bot} \to \mathcal G$ completing the top half of the pushout square defined by $\mathcal K \xrightarrow{} \mathcal L \xrightarrow{} \mathcal G$.
When the morphisms involved in the rewrite rule are monomorphisms,  as is usually required,  constructing $\mathcal{L}^{\bot}$ can be intuitively understood as finding an occurrence of $\mathcal L$ in $\mathcal G$ and removing from this occurrence everything that has no pre-image in $\mathcal K$.  The final graph $\mathcal H$ is then recovered by inserting into the hole everything from $\mathcal R$ that has no pre-image in $\mathcal K$ --- hence the two pushouts: one for deletion and one for insertion.  In this case,  we say that $\mathcal G$ is rewritten into $\mathcal H$ via the rewrite rule $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$.

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI) as shown in Fig.~\ref{fig:dpo_dpoi}~(right).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish for a rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.
By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{bonchi_string_2022-1}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.
We now make this identification throughout without further mention.
Thus we reach the following variant of DPO rewriting.
\begin{definition}[DPOI rewriting]\label{def:dpoi}
	Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the rightmost commutative diagram in Fig.~\ref{fig:dpo_dpoi} such that the two marked squares are pushouts.

	\begin{figure}[t!]
		\begin{subfigure}[T]{0.35\linewidth}
			\[
				% \tikzfig{../figures/combinatorial_semantics/DPO_square}
				\begin{tikzcd}
					{\mathcal{L}} & {\mathcal{K}} & {\mathcal{R}} \\
					{\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}}
					\arrow["f"', from=1-1, to=2-1]
					\arrow[from=1-2, to=1-1]
					\arrow[from=1-2, to=1-3]
					\arrow[from=1-2, to=2-2]
					\arrow[from=1-3, to=2-3]
					\arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
					\arrow[from=2-2, to=2-1]
					\arrow[from=2-2, to=2-3]
					\arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
				\end{tikzcd}
			\]
			% \subcaption{DPO square}
			% \label{fig:dpo}
		\end{subfigure}
		\hfill
		\begin{subfigure}[T]{0.4\linewidth}
			\[
				% \scalebox{0.8}{
				% \tikzfig{../figures/combinatorial_semantics/DPOI_square_Hyp}
				% }
				\begin{tikzcd}
					{\mathcal{L}} & {i+j} & {\mathcal{R}} \\
					{\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}} \\
					& {n+m}
					\arrow["f"', from=1-1, to=2-1]
					\arrow[from=1-2, to=1-1]
					\arrow[from=1-2, to=1-3]
					\arrow[from=1-2, to=2-2]
					\arrow[from=1-3, to=2-3]
					\arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
					\arrow[from=2-2, to=2-1]
					\arrow[from=2-2, to=2-3]
					\arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
					\arrow[from=3-2, to=2-1]
					\arrow[from=3-2, to=2-2]
					\arrow[from=3-2, to=2-3]
				\end{tikzcd}
			\]
			% \subcaption{DPOI square}
			% \label{fig:dpoi}
		\end{subfigure}
		\caption{DPO and DPOI squares}
		\label{fig:dpo_dpoi}
	\end{figure}

\end{definition}
To make DPOI rewriting sound and complete with respect to rewriting of $\Sigma$-terms modulo SMC laws and equations $\mathcal{E}$ the notion of both pushout complement and match must be restricted: otherwise there might be DPO rewrites that are not derivable from SMC laws and $\mathcal{E}$-equations.
This gives rise to the following notion of a \textit{boundary complement} and \textit{convex} DPOI rewriting.

>>>>

\section{String Diagram Rewrite Theory}%
\label{sec:combinatorial-semantics}

We recall the fundamental definitions and results of string diagram rewrite theory for symmetric monoidal categories, recapitulating the correspondence between string diagram rewriting and an appropriate notion of double pushout (DPO) rewriting of certain hypergraphs, as established in~\citet{bonchi_string_2022-1, bonchi_string_2022-2}.

First,  we fix some notation.  Let $(-)^*$ be the free monoid monad over $\catname{Set}$.
We extend this to the case of relations $R \subseteq V \times V$,  denoting by $R^{*} \subseteq V^* \times V^*$ the element-wise extension of $R$ to a relation on ordered sequences.
We will sometimes use functional notation for relations, so that given $R \subseteq V \times W$ and $v \in V$, we write $R(v) \in W$.
We elide associativity and unit isomorphisms associated with coproducts $\sqcup$,  and denote by $\iota_j\colon X_{j} \rightarrow X_{1} \sqcup \dots X_{j} \sqcup \dots X_{n}$ the $j^{\text{th}}$ injection,  and $[f,g]\colon A_1 \sqcup A_2 \to B$ the co-pairing of two morphisms $f\colon A_1 \to B$ and $g\colon A_2 \to B$.
We denote by $A \sqcup_{f,g} B$ the pushout of the span $A \xleftarrow{f} C \xrightarrow{g} B$.
We will also refer to $A,B$ as \emph{feet} of the cospan, and to $C$ as \emph{carrier}.
% Given an ordered sequence of labels from some alphabet $\Sigma$, $[l_1, \ldots, l_n]$ we will denote with $w([l_1, \ldots, l_n])$ the word concatenated from these labels using $\otimes$, e.g.\ $w([A,B]) = A \otimes B$.

@@

Hypergraphs generalise directed graphs by allowing edges to have multiple (ordered) sources and targets.
\citet{bonchi_string_2022-2} show that a certain class of hypergraphs \emph{with interfaces} and their DPO rewriting are sound and complete for categorical presentations of arbitrary symmetric monoidal theories.
Moreover, they develop the use of hypergraphs as combinatorial representations of string diagrams for symmetric monoidal categories, whereby morphisms (nodes) are represented as hyperedges, and objects (wires) as vertices.
We briefly present this correspondence, but the interested reader should consult the work of~\citet{bonchi_string_2022-1,bonchi_string_2022-2}.

\begin{definition}[Category of hypergraphs]\label{def:hypergraph}
	Over a monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$, a hypergraph $\mathcal{G}$ is a tuple $(V,E,s,t,l)$, where $V$ is a finite set of vertices, $E$ is a finite set of hyperedges, $s\colon E \to V^{*}$ is a source function, $t\colon E \to V^{*}$ is a target function,  and $l = (l_{V}\colon V \to \Sigma_{O},l_{E}\colon E \to \Sigma_{M})$ is a pair of labelling functions that assigns each hyperedge and vertex an operator and a type from monoidal signature $\Sigma$ respectively.

	The labelling function must respect the typing of the operator:
	\[
		\forall e \in E. \, l(e) = c : A \to B \implies l_{V}^{*}(s(e)) = A \text{ and } l_{V}^{*}(t(e)) = B.
	\]
	We call a hypergraph discrete if its set of hyperedges is empty.

	A \emph{hypergraph homomorphism} $\phi\colon \mathcal{F} \to \mathcal{G}$ is given by a pair of functions $\phi_V\colon V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E\colon E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that the following hold:
	\begin{align*}
		\phi_V^*(s_{\mathcal{F}}(e)) & = s_{\mathcal{G}}(\phi_E(e)),   & \phi_V^*(t_{\mathcal{F}}(e)) & = t_{\mathcal{G}}(\phi_E(e)),   \\
		l_{V,\mathcal{F}}(v)         & = l_{V,\mathcal{G}}(\phi_V(v)), & l_{E,\mathcal{F}}(e)         & = l_{E,\mathcal{G}}(\phi_E(e)).
	\end{align*}

	We denote by $\catname{Hyp(\Sigma)}$ the category of hypergraphs over $\Sigma$ and hypergraph homomorphisms.
\end{definition}
$\catname{Hyp(\Sigma)}$ is equivalently characterised as the category of (finite) presheaves over the category $\mathbf{I}$ generated by objects pairs $(k, l) \in \mathbb{N} \times \mathbb{N}$ and an additional object $\star$, with $k + l$ morphisms: $\{ i\colon \star \to (k, l) \mid 0 \leq i < k + l \}$.
As such, it has all finite colimits; the initial object is given by the empty hypergraph, and coproducts are given by the disjoint union of hypergraphs.

However, there is also a need to specify which vertices are to be considered as \emph{input} and \emph{outputs} to the diagram as a whole, corresponding to the dangling wires exiting at the top and bottom of a string diagram.
This gives rise to the definition of a hypergraph with interfaces via a construction on cospans.

% \begin{definition}[Symmetric monoidal category of cospans]
% Let $\mathbb{C}$ be a category with all finite colimits.  
% A cospan from $X$ to $Y$ is a pair of arrows $X \xrightarrow{} A \xleftarrow{} Y$  in $\mathbb{C}$.  This \emph{category of cospans of $\mathbb{C}$},  denoted $\catname{Csp(\mathbb{C})}$,  has as objects those of $\mathbb{C}$ and as arrows isomorphism classes of cospans.

% Composition of cospans $X \xrightarrow{f} A \xleftarrow{g} Y$ and $Y \xrightarrow{h} B \xleftarrow{i} Z$ is given by pushout: $X \xrightarrow{f} A \sqcup_{g,h} B \xleftarrow{i} Z$,  with identity given by the cospan of identities.  Its monoidal product is given by the coproduct of $\mathbb{C}$ as follows: 
% \begin{multline*}
%     (A \xrightarrow{f} \mathcal{G} \xleftarrow{g} B) \otimes (A^\prime \xrightarrow{f^\prime} \mathcal{G^\prime} \xleftarrow{g^\prime} B^\prime) \coloneq\\ A \sqcup A^\prime \xrightarrow{f \sqcup f^\prime} \mathcal{G} \sqcup \mathcal{G^\prime} \xleftarrow{g \sqcup g^\prime} B \sqcup B^\prime  , 
% \end{multline*}
% and its unit by the initial object of $\mathbb{C}$.
% Symmetry is inherited from $\mathbb{C}$ as the coproduct~\cite{MonoidalCoproduct}.
% \end{definition}

% Then, morphisms of $\catname{S}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces~\cite{bonchi_string_2022-2}.
\begin{definition}[Hypergraphs with interfaces]%
	\label{def:cspd}
	A hypergraph with interfaces is a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\catname{Hyp}(\Sigma)$ where $n$, $m$ are discrete hypergraphs denoting input and output interfaces.
	Write $\catname{HypI}(\Sigma)$ for the full subcategory of the quotient category of cospans\footnote{This is the 1-categorical version of the bicategory of cospans: for a bicategory, its quotient category is given by the same objects but by taking \emph{isomorphism classes} of morphisms instead.} of $\catname{Hyp}(\Sigma)$ whose objects are discrete hypergraphs and morphisms are isomorphism classes of cospans of hypergraphs\footnote{By abuse of notation, we ignore the distinction between a cospan and its isomorphism class --- this is validated by replacing \enquote{pushout} (which is only determined up to unique isomorphism) by \enquote{chosen pushouts}, which any concrete setting (e.g.\ an implementation of DPO rewriting) would satisfy.}.
\end{definition}
$\HypI{\Sigma}$ is a symmetric monoidal category:
\begin{enumerate}
	\item the tensor product given by coproduct in $\catname{Hyp}(\Sigma)$:
	      \begin{align*}
		      n \xrightarrow{f}               & \mathcal{F} \xleftarrow{g} m                                                                                                                                       \\
		                                      & \otimes \qquad \quad \Coloneqq n \sqcup n^\prime \xrightarrow{f \sqcup f^\prime} \mathcal{F} \sqcup \mathcal{G} \xleftarrow{g \sqcup g^\prime} m \sqcup m^\prime ; \\
		      n^\prime \xrightarrow{f^\prime} & \mathcal{G} \xleftarrow{g^\prime} m^\prime
	      \end{align*}
	\item the monoidal unit is given by the initial object of $\catname{Hyp}(\Sigma)$;
	\item symmetry is inherited from coproduct in $\catname{Hyp}(\Sigma)$~\cite{MonoidalCoproduct}.
\end{enumerate}

Composition of two cospans $n \xrightarrow{f} \mathcal{F} \xleftarrow{g} m;m \xrightarrow{f^\prime} \mathcal{G} \xleftarrow{g^\prime} k$ is given by the pushout
$
	n \xrightarrow{f} \mathcal{F} \sqcup_{g,f^\prime} \mathcal{G} \xleftarrow{g^\prime} k
$,
computed by identifying vertices in $\mathcal{F} \sqcup \mathcal{G}$ that share pre-image in $m$.

As it stands, every string diagram in a symmetric monoidal category can be represented by a hypergraph with interfaces, but the converse is not true.
In order to rectify this, we impose additional constraints on the hypergraphs under consideration.
\begin{definition}[Degree,  path,  acyclic]
	The in-degree of a vertex $v$ in a hypergraph $\mathcal{G}$ is the number of hyperedges $e \in {E_\mathcal{G}}$ such that $v \in t(e)$.
	Similarly, the out-degree of $v$ is the number of hyperedges $e \in E_\mathcal{G}$ such that $v \in s(e)$.
	A path from $e_0$ to $e_{n-1}$ of length $n$ in a hypergraph $\mathcal{G}$ is an ordered sequence of hyperedges $[e_0, e_1, \ldots, e_{n-1}] \in E_\mathcal{G}^*$ such that for all $i < n - 1$ some vertex $v \in t(e_i)$ is also in $s(e_{i+1})$.
	A hypergraph is directed acyclic if it contains no paths of length $n > 0$ that start and end in the same hyperedge.
\end{definition}

\begin{definition}[Category of MDA Hypergraphs with Interfaces~\cite{bonchi_string_2022-2}]%
	\label{def:monogamy_hyp}
	We call a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\HypI{\Sigma}$ monogamous if:
	\begin{enumerate}
		\item $f$ and $g$ are monomorphisms;
		\item the in-degree and out-degree of every vertex is at most~$1$;
		\item vertices with in-degree $0$ are precisely the image of $f$;
		\item vertices with out-degree $0$ are precisely the image of $g$.
	\end{enumerate}
	We denote by $\MdaCospans$ the subcategory of $\HypI{\Sigma}$ of monogamous directed acyclic (MDA) cospans.
\end{definition}

\autoref{fig:hypergraph-with-interfaces} gives an example of an MDA cospan, which is such because it \enquote{looks like} the string diagram \comonoid.
Analogously, we can see that~\autoref{fig:hypergraph} (equipped with the interfaces choosing $\{ v_1 \}$ and $\{ v_4, v_5, v_6, v_7 \}$) fails to form an MDA cospan --- it is not monogamous for many reasons: $v_1$ has out-degree 0, $v_1, v_5, v_7$ have in-degree 2, and $v_3$ has in-degree 0; additionally, it fails be be directed acyclic due to the path along $e_4$ from $v_7$ to itself\footnote{Recall that loops are not part of the syntax of string diagrams for symmetric monoidal categories in the absence of a trace --- a relaxation of monogamy to allow for this is explored by~\citet{ghica_rewriting_2023}.}.

\begin{figure}
	\begin{subfigure}[c]{0.3\linewidth}
		\centering
		\begin{tikzpicture}[
				vertex/.style={draw, circle, fill, inner sep=1.5pt},
				hyperedge/.style={draw, rectangle}
			]
			\node[vertex, label=left:{$v_4$}] (v1) at (0,0) {};
			\node[vertex, label=left:{$v_5$}] (v2) at (1,0) {};
			\node[vertex, label=left:{$v_6$}] (v3) at (2,0) {};
			\node[vertex, label=left:{$v_7$}] (v4) at (3,0) {};
			\node[vertex, label=left:{$v_2$}] (v5) at (0.5,-2) {};
			\node[vertex, label=left:{$v_3$}] (v6) at (2,-2) {};
			\node[vertex, label=left:{$v_1$}] (v7) at (2,-3) {};

			\node[hyperedge] (e1) at (0.5,-1) {$e_3$};
			\node[hyperedge] (e2) at (2,-1) {$e_4$};
			\node[hyperedge] (e3) at (0.5,-3) {$e_1$};
			\node[hyperedge] (e4) at (2,-2.5) {$e_2$};

			\draw (v1) -- (e1);
			\draw (v2) -- (e1);
			\draw (v5) -- (e1);
			\draw (v5) edge[bend right] (e3);
			\draw (v5) edge[bend left] (e3);
			\draw (v2) edge[in=130,out=-90] (e2);
			\draw (v3) -- (e2);
			\draw (v4) edge[in=50,out=-90] (e2);
			\draw (v4) edge[in=-50,out=-90] (e2);
			\draw (v6) -- (e2);
		\end{tikzpicture}
		\caption{Hypergraph with 7 vertices and 4 hyperedges, all directed bottom-to-top; source/target vertices are ordered left-to-right.}%
		\label{fig:hypergraph}
	\end{subfigure}
	\begin{subfigure}[c]{0.3\linewidth}
		\centering
		\begin{tikzpicture}[
				vertex/.style={draw, circle, fill, inner sep=1.5pt},
				hyperedge/.style={draw, rectangle}
			]
			\node[vertex, label=left:{$v_3$}] (v1) at (0,0) {};
			\node[vertex, label=left:{$v_3$}] (v1') at (0,1.5) {};
			\node[vertex, label=left:{$v_2$}] (v2) at (1,0) {};
			\node[vertex, label=left:{$v_2$}] (v2') at (1,1.5) {};
			\node[vertex, label=left:{$v_1$}] (v3) at (0.5,-2) {};
			\node[vertex, label=left:{$v_1$}] (v3') at (0.5,-3.5) {};

			\node[hyperedge] (e1) at (0.5,-1) {$e_1$};

			\draw (v1) -- (e1);
			\draw (v2) -- (e1);
			\draw (v3) -- (e1);

			\node[draw, rounded corners, fit={(v1'.west) (v2'.west)}, inner sep=12pt] (S) {};
			\node[draw, rounded corners, fit={(v3'.west)}, inner sep=12pt] (T) {};
			\node[draw, rounded corners, fit={(v1.west) (v2.west) (v3.west)}, inner sep=12pt] (H) {};
			\draw[commutative diagrams/hookrightarrow,shorten >=2pt,shorten <=2pt] (S) -- node[left] {$g$} (H);
			\draw[commutative diagrams/hookrightarrow,shorten >=2pt,shorten <=2pt] (T) -- node[left] {$f$} (H);
		\end{tikzpicture}
		\caption{Hypergraph with interfaces}%
		\label{fig:hypergraph-with-interfaces}
	\end{subfigure}
	\caption{Example hypergraphs}%
	\label{fig:hypergraph-examples}
\end{figure}

Note that $\MdaCospans$ remains a symmetric monoidal category with the same structure as before.
We now recall the following theorem~\cite{bonchi_string_2022-2}, which says there is a symmetric monoidal equivalence which witnesses that MDA hypergraphs with interfaces are precisely string diagrams.
\begin{theorem}[Corollary 26 of~\cite{bonchi_string_2022-2}]\label{thm:prop-equiv}
	$\catname{S}(\Sigma) \cong \MdaCospans$
\end{theorem}

The practical meaning of this is that two $\Sigma$-terms $f$ and $g$ are equal modulo symmetric monoidal category equations if and only if they are represented as the same hypergraph with interfaces.
The theorem defines a full and faithful functor that is induced by the interpretation function of morphism terms that renders the latter as cospans of hypergraphs and is given in~\autoref{sec:appendix:interpretation}.
Term rewriting and hence equality modulo $\mathcal{E}$ is then implemented as DPO rewriting over such hypergraphs with interfaces, which we recall next.

\subsection{DPOI-Rewriting for Hypergraphs with Interfaces}
DPO rewriting is a standard technique for formalising graph transformations categorically.
% In this section we recall standard results of double-pushout (DPO) rewriting of hypergraphs,  as well as the specific notion of DPO rewriting which is sound and complete for arbitrary SMTs,  which is called \emph{convex} rewriting.
The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $\mathcal L \rightsquigarrow \mathcal R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $\mathcal L$ within $\mathcal{G}$,  remove it,  and then plug $\mathcal R$ into the hole to determine the rewritten graph $\mathcal{H}$.
In DPO rewriting, such a rewrite rule is formalised as a span $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$ in some appropriate category of graphs, where $\mathcal{K}$ is some invariant graph fragment that should be maintained during rewriting.
A typical DPO square is depicted in~\autoref{fig:dpo_dpoi}~(left).
